Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2016-11-03 RMS/final-version/main.tex
polished all the text

It is even more crucial to ensure the reliability of an implementation
instead of the algorithm, when RMS serves in a safety-critical
system. When analyzing a realistic implementation, theoretical results
may be no more applicable. For instance, even though the conditions
derived from algorithm analysis are satisfied, schedulability can be
broken by overhead in the system, or by the interrupt mask mechanism
which may delay interrupt handling. On the other hand, correctness of
the implementation with respect to the algorithm is difficult to be
verified by the traditional methods such as testing and simulation due
to their incompleteness. Extensive effort to apply formal methods,
such as model checking and theorem proving, has been made to analyze
safety-critical systems for the past few
years~\cite{DBLP:journals/tie/MiyawakiMSYV05,DBLP:journals/iandc/MeseguerR13,DBLP:journals/cacm/Leroy09,DBLP:conf/sosp/KleinEHACDEEKNSTW09}. However,
as far as we know, few~\cite{DBLP:conf/iceccs/CuiDT14,TianD2011}
attempt to analyze the RMS algorithm, while no work for
implementations of RMS is found.

@@

These assumptions make our model different from the standard one. For
instance, with assumption~(A1'), an interrupt request that occurs
during switching will be pending, such that jobs of task $\tau_i$
cannot initiate at $kT_i$. They should wait until switching finishes
and the interrupt mask bit is cleared, which is different
from~(A1). Note that (A3) says that jobs are ready to run at their
initiation times, however, no one can start running at exactly its
initiation time, because scheduling takes time. Under these
assumptions, an example showing the execution of tasks scheduled by
the implementation is depicted in Fig.~\ref{f:example}(b). Note that
the second job instance of $\tau_1$ is initiated at time $11$ instead
of $10$, because switching is performed and the interrupt mask bit is
set at time $10$, exemplifying differences between (A1') and (A1).

@@

In this section, we introduce first how we model the states--the
static aspect--of the system using terms of given sorts, or called
data types, then how we specify the essential behaviors--the dynamic
aspect--using rewrite rules. In particular, modeling of
\emph{instantaneous behaviors} would be explained in
Sections~\ref{ss:ir}, \ref{ss:init} and~\ref{ss:inthandling}, followed
by \emph{timed behaviors} in Section~\ref{ss:timedbehavior}.

\subsection{Basic Data Types}
In our model, tasks are identified by their indexes of sort \verb|Nat|
in the $taskList$. We define a sort \verb|MaybeNat| wrapping
\verb|Nat|s to refer to some task, with \emph{constructor} \verb|some|
followed by a \verb|Nat| $n$ indicating the task indexed $n$, and
\verb|none| for no task:
\begin{verbatim}
  op none : -> MaybeNat [ctor] .
  op some_ : Nat -> MaybeNat [ctor] .
\end{verbatim}
where the keyword \verb|ctor| denotes the corresponding function
symbol to be a constructor.

A sort \verb|Stack| is introduced to model the stack of the system,
storing the tasks that are being interrupted, and equipped with
operations \verb|push|, \verb|pop| and \verb|peek| on it.
\hide{
\begin{verbatim}
  op bottom : -> Stack [ctor] .
  op _#_ : Nat Stack -> Stack [ctor] .
\end{verbatim}
}

We also need
%a sort \verb|Interval| to denote computation requirements of tasks, and 
a sort \verb|Counter| to record the execution of tasks.  We call
\emph{execution time} the time how long a task has been executing
for. A \verb|Counter| records the execution time and the computation
requirement of a task.
\hide{
\begin{verbatim}
  op [_/_] : Time Time -> Counter [ctor] .
\end{verbatim}
}

\hide{
At last, to make our model checkable by untimed model checking, it is
reasonable to reset the global variable $timer$ when it reaches an
upper bound while increasing (see Line~\ref{l:timer} in
Figure~\ref{f:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upper bound would be the least common multiple of
the periods of all tasks.}

The global variable $timer$ is reset when it reaches an upper bound
while increasing, which is not shown in detail in
Figure~\ref{f:schedule} but is reasonable. The upper bound is the
least common multiple of the periods of all tasks. A sort \verb|Timer|
is defined to model the $timer$.
\hide{
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}
}

\subsection{Modeling the System States}
The system can be considered as consisting of several parts: the tasks
which are scheduled, the scheduler itself, the hardware including
registers and stacks, and the interrupt source. The scheduler,
i.e. the function $schedule()$, can be described by a single variable
$timer$. We present the models of the other parts one by one.

@@

A periodic task, which needs to be scheduled, is an object instance of
the subclass \verb|PTask| of \verb|Task| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
%  class PTask | priority : Nat, period : NzNat, 
\begin{verbatim}
  class PTask | priority : Nat, period : Nat, 
                status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors
\verb|RUNNING|, \verb|INTERRUPT|, \verb|READY| and \verb|DORMANT|,
same as in the implementation.
\hide{
\begin{verbatim}
  ops RUNNING INTERRUPT READY DORMANT 
        : -> Status [ctor] .
\end{verbatim}
}

The list of periodic tasks, the variable $taskList$ in the
implementation, is modeled as an instance of sort \verb|TaskList|,
which is a list of \verb|PTask|s and/or \verb|error|s.
\hide{
\begin{verbatim}
  op null : -> TaskList [ctor] .
  op _::_ : Object TaskList 
              ~> TaskList [ctor] .
  mb (< O:Oid : PTask |> :: L:TaskList) 
       : TaskList .
  mb (error :: L:TaskList) : TaskList .
\end{verbatim}
}
Periodic tasks are identified by their indexes in the list.

On the other hand, a system task is an object instance of the subclass
\verb|SysTask| of \verb|Task| with no extra attributes:
\begin{verbatim}
  class SysTask | .  
  subclass SysTask < Task .
\end{verbatim}
Different from periodic tasks, system tasks are organized in a
multiset of sort \verb|SysTasks|, and identified by their \verb|Oid|s.
%We abstract here from the detailed definitions thanks to the
%similarity.

\subsubsection{Hardware}
Our model considers two parts of hardware related to the interrupt
handling mechanism: the registers and the stack.

The set of registers is modeled as an object instance of the class
\verb|Regs|, with attributes \verb|pc| denoting the program counter,
\verb|mask| for the interrupt mask bit and \verb|ir| for the interrupt
request bit, respectively:
\begin{verbatim}
  class Regs | pc : TaskID, 
               mask : Bool, ir : Bool .
\end{verbatim}
where the sort \verb|TaskID| contains subsorts \verb|MaybeNat| and
\verb|Oid|, referring to some task.
\hide{
\begin{verbatim}
  subsorts MaybeNat Oid < TaskID .
\end{verbatim}
}
Some operations, such as \verb|getPc| and \verb|setMask|, are defined
on the class \verb|Regs|.

Then the hardware is described by the sort \verb|Hardware| consisting
of an instance of \verb|Regs| and a term of sort \verb|Stack|.  
\hide{
\begin{verbatim}
  op [_;_] : Object Stack ~> Hardware [ctor] .
  mb ([ < O:Oid : Regs |> ; S:Stack ]) 
       : Hardware .
\end{verbatim}
}

@@

\subsubsection{System}
The entire system in our model is a composition of the parts
introduced above, of a sort \verb|System|\footnote{Following the Maude
  convention, variables would be written in capital letters. Some
  variable declarations are not shown for simplicity.}:
\begin{verbatim}
  op _____ : TaskList Timer SysTasks 
             Hardware Object ~> System [ctor] .
  mb (L T STS HW < O : IntSrc |>) : System .
\end{verbatim}
where ``\verb|~>|'' means that the function defined is a partial
function and the keyword \verb|mb| declares a membership axiom,
stating here that a term composed of a \verb|TaskList|, a
\verb|Timer|, a \verb|SysTasks|, a \verb|Hardware| and an object
instance of class \verb|IntSrc| is of sort \verb|System|.

\subsection{Interrupt Requests}
\label{ss:ir}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instantaneous action, thus is modeled by the
following instantaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) 
    => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the function \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
bit indicating there exists an interrupt request to be handled.
\hide{
\begin{verbatim}
  op _.intReq : Hardware -> Hardware .
  eq [ REGS ; S ].intReq 
       = [ (REGS).setIr ; S ] .
\end{verbatim}
}

Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Figure~\ref{f:schedule}, which is treated as an
instantaneous action in our model. It is modeled by a recursive
function \verb|updateStatus_with_|:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer 
                            -> TaskList . 
\end{verbatim}
%  eq updateStatus null with TIMER = null .
%  eq updateStatus (TASK :: L) with TIMER
%     = (update TASK with TIMER) 
%       :: (updateStatus L with TIMER) .
which applies function \verb|update_with_| on individual task
sequentially to update the status
(Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Figure~\ref{f:schedule}):
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, 
                           status : ST > 
        with TIMER
      = if ST == DORMANT 
        then < O : PTask | status : READY >
        else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > 
       with TIMER
     = if ST == RUNNING 
       then < O : PTask | status : INTERRUPT >
       else < O : PTask |> fi [otherwise] .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$.
Given a task, if \verb|TIMER|($timer$) can be divided by its period
\verb|T|, this task should be initiated.  In the case where the task
should be initiated, it is set \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

We can see that \verb|updateStatus_with_| behaves the same as
$updateStatus()$ in Figure~\ref{f:schedule}.

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the bit \verb|mask| to be cleared. Once the
request is detected, it is handled in two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on), and to invoke the function
$schedule()$. This behavior is modeled by the following instantaneous
rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM 
    => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The function \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i) clearing the bit \verb|ir|, which means the request
has been handled; (ii) pushing the current \verb|pc| into the stack,
storing the interrupted context; (iii) assigning \verb|scheduling| of
sort \verb|Oid| to \verb|pc|, which indicates that the system is
scheduling; and (iv) setting the bit \verb|mask|, to mask coming
interrupt requests.

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Figure~\ref{f:schedule}. It updates the status of $taskList$ and
increases $timer$ by $1$. This part is modeled by function
\verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
     = ((updateStatus L with T) 
        inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, searching the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
     = (L T (finish scheduling in STS) 
        HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Figure~\ref{f:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.

When the execution time of the system task \verb|scheduling| reaches
its computation requirement, scheduling is finished. We model this
instantaneous action with the following rule:
\begin{verbatim}
  crl [scheduling-finish] :
    (L T STS HW ISRC) 
    => (SYSTEM).finishScheduling
    if SYSTEM := (L T STS HW ISRC) 
       /\ (SYSTEM).running == scheduling 
       /\ scheduling isComplete?in STS .
\end{verbatim}
where function \verb|_.running| returns the current \verb|pc| value of
the system, and \verb|_isComplete?in_| checks whether the execution
time of the task reaches its computation requirement.  
\hide{
\begin{verbatim}
  op _mayFinish?in_ : Oid SysTasks ~> Bool .
  eq O mayFinish?in [ < O : SysTask | cnt : C > REST ] 
       = C mayFinish? .
  op _mayFinish? : Counter -> Bool .
  eq [ R / [ MIN , MAX ] ] mayFinish?
       = if R lt MIN then false else true fi .
\end{verbatim}}

Similar to scheduling, the switching stage is also divided into timed
behaviors of \verb|switching| and its functionality. \verb|switching|
starts when the running periodic task is complete, and finishes when
itself is so. Two similar instantaneous rules \verb|switching-start|
and \verb|switching-finish| are defined to model the functionality of
switching. 
\hide{
\begin{verbatim}
  crl [task-finish] :
    (L T STS HW ISRC) 
    => (SYSTEM).startSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ some N := (SYSTEM).running
       /\ some N isComplete?in L .
  crl [switching-finish] :
    (L T STS HW ISRC) 
    => (SYSTEM).finishSwitching
    if SYSTEM := (L T STS HW ISRC)
       /\ (SYSTEM).running == switching
       /\ switching isComplete?in STS .
\end{verbatim}
}

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
Timed behaviors of the system consist of two parts: the execution of
tasks and the execution of the interrupt source. Both are modeled
together by the following single \emph{standard} tick
rule\footnote{The keyword \texttt{nonexec} should be given to allow
  the Real-Time Maude engine to apply the rule with some
  strategies.}~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines the effects of time elapse on the system,
and \verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime
allowed to \emph{e}lapse from the current state until an instantaneous
transition \emph{must} be performed. In fact, the core to model timed
behaviors is to define functions \verb|delta| and \verb|mte|. Notice
that the variable \verb|R| is \emph{continuous} with respect to the
specific time domain\footnote{Real-Time Maude contains built-in
  modules to define the time domain to be natural numbers and rational
  numbers, specifying \emph{discrete} time domains and \emph{dense}
  time domains, respectively.}  that we choose to instantiate our
model on, which is different from timed automata that discretize dense
time by defining ``clock region''.

Time affects the system by advancing both the running task whose $ID$
is loaded at \verb|pc| and the interrupt source simultaneously.  While
time elapses, \verb|cnt| of the former increases and \verb|val| of the
latter decreases, respectively:
\begin{verbatim}
  ceq delta((L T STS HW ISRC), R)
      = (deltaTask(ID, L, R) 
         T STS HW (deltaIS(ISRC, R)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
where the last condition states that \verb|ID| is of sort
\verb|MaybeNat|. Due to similarity, we omit details for the case where
\verb|ID| is of sort \verb|Oid| and \verb|deltaTask| applies on
\verb|STS| instead of \verb|L|.

\verb|mte|, the maximum amount of time allowed to elapse, depends on
when the next instantaneous action must perform. Therefore, it is
decided by three arguments: the remaining time to complete the running
task, the remaining time to request the next interrupt, and whether or
not there exists an interrupt request detected for the moment:
\begin{verbatim}
  ceq mte(L T STS HW ISRC)
      = minimum(mteTask(ID, L),
                mteIS(ISRC), mteIr(HW))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
where \verb|mteIr| returns \verb|zero| if there exists an interrupt
request detected in the system, or \verb|INF| which represents
\emph{infinity} otherwise. Again we do not show the case where
\verb|ID| is of sort \verb|Oid|, which is very similar. 
\hide{
We should
point out that \verb|mteTask| computes the remaining time to reach the
maximum of the computation requirement of the task, since that is the
time at which a \verb|task-finish| transition \emph{must} happen.
}

\section{Formal Verification}
\label{s:verification}
In this section, we analyze our model of the RMS implementation within
different realistic scenarios.  Notice that from any (reasonable)
given initial state, the number of reachable states is finite, but may
be unknown, thanks to the upper bound given to $timer$, which provides
the potential for applying the untimed model checker.

\subsection{Properties}
We consider two properties in this paper: schedulability and
correctness. By schedulability, we examine whether a given task set is
schedulable by the implementation. By correctness, we verify whether
the implementation schedules periodic tasks exactly with respect to
the RMS algorithm.

>>>>

It is even more crucial to ensure the reliability of an implementation
instead of the algorithm, when RMS serves in a safety-critical
system. When analyzing a realistic implementation, theoretical results
may be no more applicable. For instance, even though the conditions
derived from algorithm analysis are satisfied, schedulability can be
broken by overhead in the system, or by the interrupt mask mechanism
which may delay interrupt handling. On the other hand, correctness of
the implementation with respect to the algorithm is difficult to be
verified by the traditional methods such as testing and simulation due
to their incompleteness. Extensive effort to apply formal methods,
such as model checking and theorem proving, has been made to analyze
safety-critical systems for the past few
years~\cite{DBLP:journals/tie/MiyawakiMSYV05,DBLP:journals/iandc/MeseguerR13,DBLP:journals/cacm/Leroy09,DBLP:conf/sosp/KleinEHACDEEKNSTW09}. However,
as far as we know, few~\cite{TianD2011,DBLP:conf/iceccs/CuiDT14}
attempt to analyze the RMS algorithm, while no work for
implementations of RMS is found.

@@

These assumptions make our model different from the standard one. For
instance, with~(A1'), an interrupt request that occurs during
switching will be pending, such that jobs of task $\tau_i$ cannot
initiate at $kT_i$. They should wait until switching finishes and the
interrupt mask bit is cleared, which is different from~(A1). (A3) says
that jobs are ready to run at their initiation times, however, no one
can start running at exactly its initiation time, because scheduling
takes time. Under these assumptions, an example showing the execution
of tasks scheduled by the implementation is depicted in
Fig.~\ref{f:example}(b). Note that the second job instance of $\tau_1$
is initiated at time $11$ instead of $10$, because switching is
performed and the interrupt mask bit is set at time $10$, exemplifying
differences between (A1') and (A1).

@@

In this section, we introduce first how we model states -- the static
aspect -- of the system using terms of given sorts, or called data
types, then how we specify essential behaviors -- the dynamic aspect
-- using rewrite rules. In particular, modeling of \emph{instantaneous
  behaviors} would be explained in Sections~\ref{ss:ir}, \ref{ss:init}
and~\ref{ss:inthandling}, followed by \emph{timed behaviors} in
Section~\ref{ss:timedbehavior}.

\subsection{Basic Data Types}
In our model, tasks are identified by their indexes of sort \verb|Nat|
in $\mathit{taskList}$. We define a sort \verb|MaybeNat| wrapping
\verb|Nat|s to refer to some task, with \emph{constructor} \verb|some|
followed by a \verb|Nat| $n$ indicating the task indexed $n$, and
\verb|none| for no task:
\begin{verbatim}
  op none : -> MaybeNat [ctor] .
  op some_ : Nat -> MaybeNat [ctor] .
\end{verbatim}
where the keyword \verb|ctor| denotes the corresponding function
symbol to be a constructor.

A sort \verb|Stack| is introduced to model the stack of the system,
storing the tasks that are being interrupted, and equipped with
operations \verb|push|, \verb|pop| and \verb|peek| on it.

We also need
%a sort \verb|Interval| to denote computation requirements of tasks, and 
a sort \verb|Counter| to record the execution of tasks.  We call
\emph{execution time} the time how long a task has been executing
for. A \verb|Counter| records the execution time and the computation
requirement of a task.

The global variable $\mathit{timer}$ is reset when it reaches an upper
bound while increasing, which is not shown in detail in
Fig.~\ref{f:schedule} but is reasonable. The upper bound is the least
common multiple of periods of all tasks. A sort \verb|Timer| is
defined to model $\mathit{timer}$.

\subsection{Modeling the System States}
The system can be considered as consisting of several parts: the tasks
which are scheduled, the scheduler itself, the hardware including
registers and stacks, and the interrupt source. The scheduler,
i.e.\ the function $\mathit{schedule()}$, can be described by a single
variable $\mathit{timer}$. We present the models of the other parts
one by one.

@@

A periodic task, which needs to be scheduled, is an object instance of
the subclass \verb|PTask| of \verb|Task| with additional attributes
\verb|priority|, \verb|period| and \verb|status|:
\begin{verbatim}
  class PTask | priority : Nat, period : Nat, 
                status : Status .
  subclass PTask < Task .
\end{verbatim}
where \verb|Status| is a sort with four constant constructors
\verb|RUNNING|, \verb|INTERRUPT|, \verb|READY| and \verb|DORMANT|,
same as in the implementation.

The list of periodic tasks, the variable $\mathit{taskList}$ in the
implementation, is modeled as an instance of sort \verb|TaskList|,
which is a list of \verb|PTask|s and/or \verb|error|s.  Periodic tasks
are identified by their indexes in the list.

On the other hand, a system task is an object instance of the subclass
\verb|SysTask| of \verb|Task| with no extra attributes.
\hide{
\begin{verbatim}
  class SysTask | .  
  subclass SysTask < Task .
\end{verbatim}
}
Different from periodic tasks, system tasks are organized in a
multiset of sort \verb|SysTasks|, and identified by their \verb|Oid|s.

\subsubsection{Hardware}
Our model considers two parts of hardware related to interrupt
handling: the registers and the stack.

The set of registers is modeled as an object instance of class
\verb|Regs|, with attributes \verb|pc| denoting the program counter,
\verb|mask| for the interrupt mask bit and \verb|ir| for the interrupt
request bit:
\begin{verbatim}
  class Regs | pc : TaskID, 
               mask : Bool, ir : Bool .
\end{verbatim}
where the sort \verb|TaskID| contains subsorts \verb|MaybeNat| and
\verb|Oid|, referring to some task. Some operations, such as
\verb|getPc| and \verb|setMask|, are defined on the class \verb|Regs|.

Then the hardware is described by the sort \verb|Hardware| consisting
of an instance of \verb|Regs| and a term of sort \verb|Stack|.

@@

\subsubsection{System}
The entire system in our model is a composition of the parts
introduced above, of a sort \verb|System|\footnote{Following the Maude
  convention, variables would be written in capital letters. Variable
  declarations are not shown for simplicity.}:
\begin{verbatim}
  op _____ : TaskList Timer SysTasks 
             Hardware Object ~> System [ctor] .
  mb (L T STS HW < O : IntSrc |>) : System .
\end{verbatim}
where ``\verb|~>|'' means that the function defined is a partial
function and the keyword \verb|mb| declares a membership axiom,
stating here that a term composed of a \verb|TaskList|, a
\verb|Timer|, a \verb|SysTasks|, a \verb|Hardware| and an object
instance of class \verb|IntSrc| is of sort \verb|System|.

\subsection{Interrupt Requests}
\label{ss:ir}
Interrupt requests are performed by the source exactly every cycle
$T$, when the attribute \verb|val| decreases to \verb|zero|. The
requesting is an instantaneous action, thus is modeled by the
following instantaneous conditional rule applied on \verb|System|:
\begin{verbatim}
  crl [interrupt-request] :
    (L T STS HW ISRC) 
    => (L T STS (HW).intReq reset(ISRC))
    if (ISRC).timeout .
\end{verbatim}
where the function \verb|_.timeout| examines whether the attribute
\verb|val| equals \verb|zero|, and \verb|_.intReq| sets the \verb|ir|
bit indicating there exists an interrupt request to be handled.
Then the request will wait to be handled, which is explained in
Section~\ref{ss:inthandling}.

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by $\mathit{updateStatus()}$
in Fig.~\ref{f:schedule}, which is treated as an instantaneous action
in our model. It is modeled by function \verb|updateStatus_with_|:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer 
                            -> TaskList . 
\end{verbatim}
which applies function \verb|update_with_| on individual task
sequentially to update the status (Lines~21--30 in
Fig.~\ref{f:schedule}):
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, 
                           status : ST > 
        with TIMER
      = if ST == DORMANT 
        then < O : PTask | status : READY >
        else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > 
       with TIMER
     = if ST == RUNNING 
       then < O : PTask | status : INTERRUPT >
       else < O : PTask |> fi [otherwise] .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $\mathit{timer}$.
Given a task, if \verb|TIMER|($\mathit{timer}$) can be divided by its period
\verb|T|, this task should be initiated.  In the case where the task
should be initiated, it is set \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

We can see that \verb|updateStatus_with_| behaves the same as
$\mathit{updateStatus()}$ in Fig.~\ref{f:schedule}.

\subsection{Interrupt Handling and Task Scheduling}
\label{ss:inthandling}
When an interrupt request occurs, it may not be detected immediately
by the system. It requires the bit \verb|mask| to be cleared. Once the
request is detected, it is handled in two steps: the interrupt
handling mechanism of the hardware (such as clearing \verb|ir|,
pushing context into stack and so on), and to invoke the function
$\mathit{schedule()}$. This behavior is modeled by the following instantaneous
rewrite rule:
\begin{verbatim}
  crl [interrupt-handle] :
    SYSTEM 
    => ((SYSTEM).interrupt).startScheduling
    if (SYSTEM).existInt .
\end{verbatim}
where \verb|_.existInt| checks whether \verb|mask| is cleared
\emph{and} \verb|ir| is set. The function \verb|_.interrupt| models
the interrupt handling mechanism performed by the hardware and does
four things: (i)~clearing the bit \verb|ir|, which means the request
has been handled; (ii)~pushing the current \verb|pc| into the stack,
storing the interrupted context; (iii)~assigning \verb|scheduling| of
sort \verb|Oid| to \verb|pc|, which indicates that the system is
scheduling; and (iv)~setting the bit \verb|mask|, to mask coming
interrupt requests.

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~3--4 in Fig.~\ref{f:schedule}. It updates the
status of $\mathit{taskList}$ and increases $\mathit{timer}$ by
$1$. This part is modeled by function \verb|_.startScheduling| which
applies instantaneously at the beginning of scheduling, as shown in
rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
     = ((updateStatus L with T) 
        inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to Lines~6--11, searching the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
     = (L T (finish scheduling in STS) 
        HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models Lines~6--11,
searching the task with highest priority that has status
\verb|INTERRUPT| or \verb|READY| then performing an \emph{interrupt
  return} or executing it, respectively.

When the execution time of the system task \verb|scheduling| reaches
its computation requirement, scheduling is finished. We model this
instantaneous action with the following rule:
\begin{verbatim}
  crl [scheduling-finish] :
    (L T STS HW ISRC) 
    => (SYSTEM).finishScheduling
    if SYSTEM := (L T STS HW ISRC) 
       /\ (SYSTEM).running == scheduling 
       /\ scheduling isComplete?in STS .
\end{verbatim}
where function \verb|_.running| returns the current \verb|pc| value of
the system, and \verb|_isComplete?in_| checks whether the execution
time of the task reaches its computation requirement.

Similar to scheduling, the switching stage is also divided into timed
behaviors of \verb|switching| and its functionality. \verb|switching|
starts when the running periodic task is complete, and finishes when
itself is so. Two similar instantaneous rules \verb|switching-start|
and \verb|switching-finish| are defined to model the functionality of
switching.

\subsection{Timed Behaviors of the System}
\label{ss:timedbehavior}
Timed behaviors of the system consist of two parts: the execution of
tasks and the execution of the interrupt source. Both are modeled
together by the  single \emph{standard} tick
rule\footnote{The keyword \texttt{nonexec} should be given to allow
  the Real-Time Maude engine to apply the rule with some
  strategies.}~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{verbatim}
  crl [tick]:
    {SYSTEM} => {delta(SYSTEM, R)} in time R 
    if R le mte(SYSTEM) [nonexec] .
\end{verbatim}
where \verb|delta| defines effects of time elapse on the system, and
\verb|mte| denotes the \emph{m}aximum amount of \emph{t}ime allowed to
\emph{e}lapse from the current state until an instantaneous action
\emph{must} happen. In fact, the key to modeling timed behaviors is to
define \verb|delta| and \verb|mte|. Note that the variable \verb|R| is
\emph{continuous} with respect to the specific time
domain\footnote{Real-Time Maude contains built-in modules to define
  the time domain to be natural numbers and rational numbers,
  specifying \emph{discrete} time domains and \emph{dense} time
  domains, respectively.}  that we choose to instantiate our model on,
which is different from timed automata that discretize dense time by
defining ``clock region''.

Time affects the system by advancing both the running task, whose
$\mathit{ID}$ is loaded at \verb|pc|, and the interrupt source
simultaneously.  While time elapses, \verb|cnt| of the former
increases and \verb|val| of the latter decreases, respectively:
\begin{verbatim}
  ceq delta((L T STS HW ISRC), R)
      = (deltaTask(ID, L, R) 
         T STS HW (deltaIS(ISRC, R)))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
where the last condition indicates that \verb|ID| is of sort
\verb|MaybeNat|. Due to similarity, we omit details for the case where
\verb|ID| is of sort \verb|Oid| and \verb|deltaTask| applies on
\verb|STS| instead of \verb|L|.

\verb|mte| depends on when the next instantaneous transition must
perform. Therefore, it is decided by three arguments: the remaining
time to complete the running task, the remaining time to request the
next interrupt, and whether or not there exists an interrupt request
detected for the moment:
\begin{verbatim}
  ceq mte(L T STS HW ISRC)
      = minimum(mteTask(ID, L),
                mteIS(ISRC), mteIr(HW))
      if ID := (HW).getPc /\ ID :: MaybeNat .
\end{verbatim}
where \verb|mteIr| returns \verb|zero| if there exists an interrupt
request detected in the system, or \verb|INF| which represents
\emph{infinity} otherwise. The case where \verb|ID| is of sort
\verb|Oid| is similar.

\section{Formal Verification}
\label{s:verification}
In this section, we analyze our model of the RMS implementation within
different realistic scenarios.  Notice that from any (reasonable)
given initial state, the number of reachable states is finite, but may
be unknown, thanks to the upper bound given to $\mathit{timer}$, which
provides the potential for applying the untimed model checker.

\subsection{Properties}
We take two properties into account in this paper: schedulability and
correctness. By schedulability, we examine whether a given task set is
schedulable by the implementation. By correctness, we verify whether
the implementation schedules periodic tasks exactly with respect to
the RMS algorithm.

To verify the schedulability of a given set of periodic tasks, we
define an atomic proposition \verb|taskTimeout| to hold if there
exists an \verb|error| in $\mathit{taskList}$ of the current state,
that is, some task misses its deadline:
\begin{verbatim}
  op taskTimeout : -> Prop [ctor] .
  eq {L T STS HW ISRC} |= taskTimeout 
     = containError(L) .
\end{verbatim}
where \verb|containError| returns \verb|true| if there is an
\verb|error| existing in \verb|L|. Then schedulability can be
formalized as the temporal logic formula: \verb|[](~taskTimeout)|,
expressing that the proposition \verb|taskTimeout| is always false. As
the property is not \emph{clock-related}, given an initial state
\verb|init|, the following untimed model checking command returns
\verb|true| if the schedulability property holds with no time limit;
otherwise a trace showing a counterexample is provided:
\begin{verbatim}
  (mc init |=u [](~taskTimeout) .)
\end{verbatim}
